\texttt{compGeometeR}: an \texttt{R} package for computational geometry

### 2021-03-10 paper/main.tex
Updated paper

\usepackage{natbib}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref} % to insert a web link
\usepackage{amsfonts} % to get blackboard font

@@

Computational geometry seeks to develop algorithms and data structures for geometric objects, and the results are widely applied across numerous scientific domains \citep{de-berg-2008}.  Given the importance of computational geometry it is no surprise that within the R computing ecosystem there are a variety of options for applying various computational geometry algorithms in a variety of dimensions.  However, these packages often work in specific numbers of dimensions, have incompatible data structures, and include additional non-computational geometry functionality that can be domain specific.

Our objective in developing the \texttt{compGeometeR} package is to produce a systematic implementation of some of the most commonly used computational geometry algorithms across various of branches of geometry so that they can be easily integrated into domain specific scientific workflows.  The functionality of \texttt{compGeometeR} is entirely dependent on computational foundations provided by the \texttt{Qhull} C++ interface software \citep{barber-1996}.

@@

@@

@@

A convex hull \citep{barber-1996} defines the smallest subset of $\mathbb{R}^n$ that contains $P$ and for which the subset is convex so any two points within the convex hull can be connected by a straight line also contained by the convex hull (Figure \ref{fig:discrete-algorithms}a).

@@

\begin{figure}[ht]
\centering
\includegraphics{figures/figure-1/figure-1.png}
\caption{Two-dimensional examples of discrete geometry algorithms currently available in \texttt{compGeometeR}. (a) Convex hull, (b) convex layers, (c) Delaunay triangulation, and (d) alpha complex.}
\label{fig:discrete-algorithms}
\end{figure}

@@

Digital geometry is a branch of geometry that studies the geometric properties of a grid (or lattice) of points in Euclidean space, and usually involves the digitisation of discrete lines and regions \citep{rosenfeld-1989}.  Digitisation occurs by representing Euclidean space $\mathbb{R}^n$ as a a rectangular orthogonal grid $\mathbb{G}^n$.  The elements of $\mathbb{G}^2$ are called pixels, and the elements of $\mathbb{G}^3$ are called voxels, and each element in $\mathbb{G}^n$ has a grid coordinate for its centre \citep{klette-2004} and a value that denotes if the element belongs to a digitised geometric object, or when required, which part of the digitised geometric object.  \texttt{compGeometeR} has implemented digital versions of the convex hull (Figure \ref{fig:digital-algorithms}a), alpha complex (Figure \ref{fig:digital-algorithms}b), and alpha shape (Figure \ref{fig:digital-algorithms}c).

\begin{figure}[ht]
\centering
\includegraphics{figures/figure-2/figure-2.png}
\caption{Two-dimensional examples of digital geometry algorithms currently available in \texttt{compGeometeR}. (a) Convex hull, (b) alpha complex, and (c) alpha shape.}
\label{fig:digital-algorithms}
\end{figure}

@@

\section{Code availability}

@@

>>>>

\usepackage{natbib}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref} % to insert a web link
\usepackage{amsfonts} % to get blackboard font

\usepackage{listings} % to get code blocks
\usepackage{xcolor}
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.9,0.9,0.9}
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}
\lstset{style=mystyle}

@@

Computational geometry seeks to develop algorithms and data structures for geometric objects, and the results are widely applied across numerous scientific domains \citep{de-berg-2008}.  Given the importance of computational geometry it is no surprise that within the \texttt{R} computing ecosystem there are a variety of options for applying various computational geometry algorithms.  However, these packages often work in specific numbers of dimensions, have incompatible data structures, and include additional non-computational geometry functionality that can be domain specific.

Our objective in developing the \texttt{compGeometeR} package is to produce a systematic implementation of some of the most commonly used computational geometry algorithms across various of branches of geometry so that they can be easily integrated into domain specific scientific workflows.

@@

The functions of \texttt{compGeometeR} interconnect and build upon one another, but it is important to recognise that virtually all of the \texttt{compGeometeR} functions are  dependent on computational geometry foundations provided by the \texttt{Qhull} C++ interface software \citep{barber-1996}.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=15cm]{figures/software-structure/software-structure.png}
\caption{Connections and dependencies of the \texttt{compGeometeR} functions.}
\label{fig:software-structure}
\end{figure}

@@

A convex hull \citep{barber-1996} defines the smallest subset of $\mathbb{R}^n$ that contains $P$ and for which the subset is convex so any two points within the convex hull can be connected by a straight line also contained by the convex hull (Figure \ref{fig:discrete-algorithms}a).  As an example of the simplicity of \texttt{compGeometeR} a convex hull can be generated and visualised with minimal code (Listing \ref{code:discrete-convex-hull-code}).

\begin{lstlisting}[language=R, caption=Example \texttt{R} code to create a discrete convex hull with \texttt{compGeometeR}, label={code:discrete-convex-hull-code}]
library(compGeometeR)
# Generate point data
set.seed(2) # to reproduce figure exactly
x = rgamma(n = 20, shape = 3, scale = 2)
y = rnorm(n = 20, mean = 10, sd = 2)
p = cbind(x, y)
# Create convex hull
ch = convex_hull(p)
# Plot point data and convex hull
plot(x, y, yaxt="n", xaxt="n", xlab="", ylab = "", pch=16, cex=0.75)
polygon(ch$hull_vertices, col="orange", border="firebrick")
points(p, pch=16, cex=0.75)
\end{lstlisting}

@@

\begin{figure}[ht]
\centering
\includegraphics{figures/discrete-algorithms/discrete-algorithms.png}
\caption{Two-dimensional examples of discrete geometry algorithms currently available in \texttt{compGeometeR}. (a) Convex hull, (b) convex layers, (c) Delaunay triangulation, and (d) alpha complex.}
\label{fig:discrete-algorithms}
\end{figure}

@@

Digital geometry is a branch of geometry that studies the geometric properties of a grid (or lattice) of points in Euclidean space, and usually involves the digitisation of discrete lines and regions \citep{rosenfeld-1989}.  Digitisation occurs by representing Euclidean space $\mathbb{R}^n$ as a a rectangular orthogonal grid $\mathbb{G}^n$.  The elements of $\mathbb{G}^2$ are called pixels, and the elements of $\mathbb{G}^3$ are called voxels, and each element in $\mathbb{G}^n$ has a grid coordinate for its centre \citep{klette-2004} and a value that denotes if the element belongs to a digitised geometric object, or when required, which part of the digitised geometric object.  \texttt{compGeometeR} has implemented digital versions of the convex hull (Figure \ref{fig:digital-algorithms}a), alpha complex (Figure \ref{fig:digital-algorithms}b), and alpha shape (Figure \ref{fig:digital-algorithms}c) that are unique functions amongst \texttt{R} software (Table \ref{tab:r-options}).

The code required for digital versions of the discrete algorithms is no more complicated, and simply requires addtional parameters to define the extent and resolution of $\mathbb{G}^n$.  For example, the code required for a digital convex hull (Listing \ref{code:digital-convex-hull-code}) does not differ much from that required for a discrete convex hull (Listing \ref{code:discrete-convex-hull-code}).

\begin{lstlisting}[language=R, caption=Example \texttt{R} code to create a digital convex hull with \texttt{compGeometeR}, label={code:digital-convex-hull-code}]
library(compGeometeR)
# Generate point data
set.seed(2) # to reproduce figure exactly
x = rgamma(n = 20, shape = 3, scale = 2)
y = rnorm(n = 20, mean = 10, sd = 2)
p = cbind(x, y)
# Create digital convex hull
d_ch = digital_convex_hull(p, mins=c(0,5), maxs=c(15,15), spacings = c(0.05,0.05))
# Plot point data and digital convex hull
image(x=d_ch[[3]][[1]], y=d_ch[[3]][[2]], z=d_ch[[1]], 
      yaxt="n", xaxt="n", xlab="", ylab = "", col=c("white", "orange"))
points(p, pch=16, cex=0.75)
\end{lstlisting}

\begin{figure}[ht]
\centering
\includegraphics{figures/digital-algorithms/digital-algorithms.png}
\caption{Two-dimensional examples of digital geometry algorithms currently available in \texttt{compGeometeR}. (a) Convex hull, (b) alpha complex, and (c) alpha shape.}
\label{fig:digital-algorithms}
\end{figure}

@@

\section{Software availability}

@@

We are also developing a cookbook of examples of \texttt{compGeometeR} use via the GitHub repository wiki \url{https://github.com/manaakiwhenua/compGeometeR/wiki}.
